Final Project Part 3 Report

### 2024-12-06 final_paper/final_report.Rmd
commit

-   **Gadiel David Flores:** Methods, flowchart and Results
-   **Wendy Huang:** Description of contributions.

@@

Based on prior experience, we hypothesize that housing supply is a primary determinant of house prices. Therefore, we selected key indicators of housing supply as predictors: the number of available homes, homes sold, homes under construction, new homes completed, and the rate hike, which serves as a proxy for investment conditions. This analysis aims to use the multiple linear regression model to investigate the relationship between house prices and these factors, with the rate hike modeled as a categorical variable capture the impact of interest rate changes on housing affordability.

@@

In the context of rate hikes, the CEPR report discusses their inclusion in econometric models to study the effects of borrowing costs on housing affordability and prices. It finds that higher rates reduce demand, reflected in lower prices, a relationship frequently captured through regression techniques (@whatdriveshouseprices)

Our literature review confirms that the selected predictors—houses sold, houses constructed, houses under construction, unsold houses, and rate hikes—have been previously studied and exhibit a linear relationship with house prices. Linear regression is recognized as a valid method for this analysis, offering a clear way to quantify the relationships between house prices (the response) and these predictors. The validity of the linear relationship will be tested and thoroughly discussed later, as this is a key objective of this paper.

For this analysis, the focus will be on identifying and validating the linear relationship between the predictors and the response variable. We will employ exploratory data analysis (EDA) and evaluate model performance using metrics such as AIC, BIC, and adjusted $R^2$. Through these statistical methods, we aim to establish that a linear regression relationship exists between house prices and the predictors: the number of available homes, homes sold, homes under construction, homes completed, and rate hikes.

The structure of this paper is as follows: In the Methods section, we describe the dataset acquisition process and the data cleaning steps undertaken for subsequent analysis. Result section introduce the result of the linear regression and the validation of the model. Limitation section discuss the limitation of the model and dataset selection and processing. Ethics are discussed in the ethics section.

@@

To prepare the data for analysis, we used `tidyverse` for data manipulation and `car` for diagnostics. The New Housing Price Index (NHPI) was defined as the response variable, with 13 predictors identified, including absorption rates, GDP, CPI, and construction statistics. The dataset was cleaned by removing missing values, duplicate observations, and influential outliers identified using Cook's Distance. Finally, the data was randomly split into training (50%) and testing (50%) sets to ensure unbiased and robust analysis.

@@

# Results {#sec-results}

This section presents the results of the multiple linear regression (MLR) analysis. The response variable for this analysis was the New Housing Price Index (NHPI), and eight predictors were selected to represent factors such as absorption, interest rates, GDP, CPI, and various housing construction metrics. These predictors were chosen based on their theoretical relevance to housing price trends and their potential statistical significance. The focus of this model is explanatory, as the aim is to understand the relationships between variables.

@@

# Ethics Discussion {#sec-ethics}

The dataset used in this analysis originates from Statistics Canada, a reputable and reliable source, which ensures the data’s credibility. (StatCan) However, the sensitivity of the response variable—house prices—requires careful ethical consideration. Housing affordability impacts individuals and communities significantly, and any conclusions drawn from this analysis may influence public perception or policy decisions.

First, transparency is essential. We document every step of data cleaning, analysis, and model selection to ensure reproducibility and avoid misrepresentation of findings. This transparency fosters trust and helps prevent the ethical pitfalls of obscuring methodology. We are fully aware that the results are closely tied to regional economic factors and GDP. We took extra care to avoid errors that could lead to incorrect conclusions.

@@

Third, fairness in data representation is critical. The predictors—such as sold and unsold houses, construction activity, and rate hikes were analyzed impartially to avoid introducing biases. Any biases identified in the dataset or methodology must be explicitly addressed to maintain the analysis’s integrity.

>>>>

-   **Gadiel David Flores:** Methods, flowchart and Results
-   **Wendy Huang:** Introduction and Ethnic Discussion

@@

Based on prior experience, we hypothesize that housing demand, supply, and inflation are key determinants of house prices. To test this hypothesis, we identified critical indicators representing these factors: house absorption, construction activity, and the Consumer Price Index (CPI), which serves as a proxy for investment conditions. We sourced dataset with relevant predictors from Statistics Canada and conducted an in-depth analysis of the variables. After cleaning the data and assessing multicollinearity using the Variance Inflation Factor (VIF), we refined our predictors to include CPI, unabsorbed homes, and completed construction homes as the most relevant variables.This study employs a multiple linear regression model to examine the relationship between house prices and these predictors, providing insights into the factors that drive housing market dynamics.

@@

In the context of CPI, the CEPR report discusses their inclusion in econometric models to study the effects of borrowing costs on housing affordability and prices. It finds that higher CPI indicates higher inflation, reflected in higher mortgage rates and reducing both housing demand and price growth, a relationship frequently captured through regression techniques (@whatdriveshouseprices)

Our literature review confirms that the selected predictors—CPI, unabsorbed homes, and completed construction homes—have been previously studied and exhibit a linear relationship with house prices. Linear regression is recognized as a valid method for this analysis, offering a clear way to quantify the relationships between house price (the response) and these predictors. The validity of the linear relationship will be tested and thoroughly discussed later, as this is a key objective of this paper.

For this analysis, the focus will be on identifying and validating the linear relationship between the predictors and the response variable. We will employ exploratory data analysis (EDA) and Variance Inflation Factor (VIF), evaluate model performance using metrics such as AIC, BIC, $R^2$, and adjusted $R^2$. Through these statistical methods, we aim to establish that a linear regression relationship exists between house prices and the predictors: CPI, unabsorbed homes, and completed construction homes.

The structure of this paper is as follows: Methods section outlines the primary data preparation process, detailing how we refined the dataset predictors to identify the most relevant variables using exploratory data analysis (EDA) and Variance Inflation Factor (VIF) testing. Additionally, we describe the steps involved in model development, including building, diagnosing, evaluating performance, and validating the multiple linear regression model. In result section, we present the performance of the multiple linear regression model based on the selected predictors, highlighting key findings and insights. Limitation section discusses the limitations of the study, including constraints in the dataset selection, processing, and the model's scope. Ethical considerations related to the study, including data usage and implications, are addressed in Ethics section.

@@

According to the topic of house prices and its influencers, we download the dataset with relevant predictors from Statistics Canada (Statistics Canada, 2024). The New Housing Price Index (NHPI) was defined as the response variable, with 13 predictors identified, including absorption (detached-absorption, semi-absorption, detached-unabsorbed, semi-unabsorbed), GDP, CPI, and construction (starting-detached-construction, starting-semi-construction, under construction detached, under construction-semi, completed-construction-detached, completed-construction-semi).

To prepare the data for analysis, we used `tidyverse` for data manipulation and `car` for diagnostics. The dataset was cleaned by removing missing values, duplicate observations, and influential outliers identified using Cook's Distance. Finally, the data was randomly split into training (50%) and testing (50%) sets to ensure unbiased and robust analysis.

@@

# Results

This section presents the results of the multiple linear regression (MLR) analysis. The response variable for this analysis was the New Housing Price Index (NHPI), and eight predictors were selected to represent factors such as absorption, interest rates, GDP, CPI, and various housing construction metrics(). These predictors were chosen based on their theoretical relevance to housing price trends and their potential statistical significance. The focus of this model is explanatory, as the aim is to understand the relationships between variables.

@@

# Ethics Discussion

The dataset used in this analysis originates from Statistics Canada, a reputable and reliable source, which ensures the data’s credibility (Statistics Canada, 2024). However, the sensitivity of the response variable—house prices—requires careful ethical consideration. Housing affordability impacts individuals and communities significantly, and any conclusions drawn from this analysis may influence public perception or policy decisions.

First, transparency is essential. We document every step of data cleaning, analysis, and model selection to ensure reproducibility and avoid misrepresentation of findings. This transparency fosters trust and helps prevent the ethical pitfalls of obscuring methodology. We are fully aware that the results are closely tied to regional economic factors like GDP and CPI. We took extra care to avoid errors that could lead to incorrect conclusions.

@@

Third, fairness in data representation is critical. The predictors—such as sold and unsold houses, construction activity, and CPI were analyzed impartially to avoid introducing biases. Any biases identified in the dataset or methodology must be explicitly addressed to maintain the analysis’s integrity.
